BIOST 578 HW 2
========================================================

We are asked to reproduce the results presented in Figure 2 of this paper: Qian, F., Bolen, C. R., Jing, C., Wang, X., Zheng, W., Zhao, H., et al. (2013). Impaired toll-like receptor 3-mediated immune responses from macrophages of patients chronically infected with hepatitis C virus. Clinical and Vaccine Immunology : CVI, 20(2), 146–155. doi:10.1128/CVI.00530-12
 
# Get the data from GEO

```{r, eval=FALSE}
## install packages
source("http://bioconductor.org/biocLite.R")
biocLite(c("GEOquery", "lumi", "limma"))
```

```{r, message=FALSE}
## load the GEOquery package
library(GEOquery)
```
 
```{r, eval=FALSE}
## download data 
gds <- getGEO("GSE40812", destdir="/Users/jiajin/Desktop/Winter2014/BIOST578/HW/HW2-jjkee")
```
The data was downloaded in the compressed (.gz) format. I manually uncompressed the data by clicking on its icon. 

```{r}
## read the data into R 
gds <- getGEO(filename = "/Users/jiajin/Desktop/Winter2014/BIOST578/HW/HW2-jjkee/GSE40812_series_matrix.txt")
```

```{r}
## function to sanitize data
sanitize_pdata <- function(pd) {
  keepCols <- c("title", "characteristics_ch1", "characteristics_ch1.1", "characteristics_ch1.2")
  pd <- pd[, keepCols]
  colnames(pd) <- c("ptid", "infection.status", "cell.type", "treatment")
  pd$ptid <- sapply(pd$ptid, FUN=function(x) {unlist(strsplit(as.character(x), "_"))[2]})  
  pd$infection.status <- tolower(gsub("infection status: ", "", pd$infection.status))  ## neg/pos 
  pd$cell.type <- tolower(gsub("cell type: ", "", pd$cell.type))  ## monocyte-derived macrophage/pbmc
  pd$treatment <- tolower(gsub("treatment: ", "", pd$treatment))  ## poly ic h/mock 
  return(pd)
}

## keep a copy of the original data
gds_original <- gds   

## sanitize data
pData(gds) <- sanitize_pdata(pData(gds))
```

# Normalize the data

```{r, message=FALSE}
## load the lumi package
library(lumi)
```

```{r}
## normalize data 
gds <- lumiN(gds)
```
Note that the expression matrix has actually been quantile normalized using the lumi package. I am doing this step for practice; performing the normalization step again will not change the data. 

# Use limma to test for differential expression

Note that we will only be looking at macrophage samples. This step is comprised of 2 parts. First we have to find the subset of genes (or probes) that are responsive to poly-IC treatment. Secondly, among the subset of genes, find those that are differentially expressed between VL-/VL+ samples.

Get only the macrophage samples from data and sort them by treatment (poly-IC vs mock), infection status (VL- vs VL+) and patient ID. 

```{r}
macrophage.data <- gds[, pData(gds)$cell.type=="monocyte-derived macrophage"]
data.order <- with(pData(macrophage.data), order(treatment, infection.status, ptid))
pData(macrophage.data) <- pData(macrophage.data)[data.order, ]
exprs(macrophage.data) <- exprs(macrophage.data)[, data.order]
```

Find the subset of probes (or genes) that are responsive to poly-IC treatment with FDR cutoff of 0.05 and fold change of >1.5.

```{r, message=FALSE}
# load the limma package
library(limma)
```

```{r}
## test for differential expression using limma
design1 <- model.matrix(~treatment, macrophage.data)
fit1 <- lmFit(macrophage.data, design1)
ebay1 <- eBayes(fit1)

## find differentially expressed genes (or probes) with FDR cutoff of 0.05 and fold change of >1.5
topTable1 <- topTable(ebay1, coef="treatmentpoly ic h", number=Inf, p.value=0.05, lfc=log2(1.5), sort.by="p")
topProbes1 <- topTable1$Probe_Id

## check if the number of probes and genes reported in the paper the same as what I found
## the paper reported they found 1146 poly-IC responsive probes (753 upregulated in poly-IC samples, 393 downregulated in poly-IC samples), which corresponds to 977 poly-IC responsive genes (622 upregulated in poly-IC samples, 355 downregulated in poly-IC samples)
length(topProbes1)  ## number of responsive probes
sum(topTable1$logFC > 0)  ## number of upregulated probes
sum(topTable1$logFC < 0)  ## number of downregulated probes
length(unique(topTable1$Symbol)[unique(topTable1$Symbol) != ""])  ## number of responsive genes; note there is one gene without a symbol name, need to tell R not to count it
logFCgtr0 <- unique(topTable1$Symbol[which(topTable1$logFC > 0)]); length(logFCgtr0[logFCgtr0 != ""])  ## number of upregulated genes
logFCless0 <- unique(topTable1$Symbol[which(topTable1$logFC < 0)]); length(logFCless0[logFCless0 != ""])  ## number of downregulated genes
```

For each probe found in previous step, we need to first calculate fold change between each subject's paired mock and poly-IC sample.

```{r}
## select the appropriate subset of data for further analysis
subset.data <- macrophage.data[rownames(exprs(macrophage.data)) %in% topProbes1, ]

## construct matrix to multiply by the expression value matrix such that values that correspond to "mock" samples have a value of 1 and "poly ic h" samples have a value of -1
treatment.matrix <- matrix(0, nrow=nrow(exprs(subset.data)), ncol=ncol(exprs(subset.data)))
treatment.matrix[, which(pData(subset.data)$treatment=="mock")] <- 1
treatment.matrix[, which(pData(subset.data)$treatment=="poly ic h")] <- -1

## construct matrix to identify samples belonging to each patient
patient.matrix <- model.matrix(~ptid, subset.data)

## modify patient matrix so that the first column corresponds to the first patient
first.ptid <- pData(subset.data)$ptid[1]
colnames(patient.matrix)[1] <- paste0("ptid", first.ptid)
patient.matrix[which(pData(subset.data)$ptid != first.ptid), 1] <- 0 

## calculate fold change between each subject's paired mock and poly-IC sample
exprs(subset.data) <- (exprs(subset.data) * treatment.matrix) %*% patient.matrix

## limit data to one row per patient (just choose the one corresponding to poly-IC)
pData(subset.data) <- pData(subset.data)[pData(subset.data)$treatment=="poly ic h", ]  
```

Now fold change values for VL- patients were compared with those of VL+ patients using limma.

```{r}
## test for differential expression using limma
design2 <- model.matrix(~infection.status, subset.data)
fit2 <- lmFit(subset.data, design2)
ebay2 <- eBayes(fit2)

## find differentially expressed genes (or probes) with unadjusted p-value < 0.1
topTable2 <- topTable(ebay2, coef="infection.statuspos", adjust.method="none", number=Inf, p.value=0.1)  
topProbes2 <- topTable2$Probe_Id

## check if the number of probes and genes reported in the paper the same as what I found
## the paper reported they found 43 differentially expressed probes (30 upregulated in VL- samples, 13 upregulated in VL+ samples)
length(topProbes2)  ## number of probes
sum(topTable2$logFC > 0)  ## number of upregulated probes in VL+ samples
sum(topTable2$logFC < 0)  ## number of upregulated probes in VL- samples
```

# Display the results using a heatmap

```{r, eval=FALSE}
## install package
install.packages("gplots")
```

```{r, message=FALSE}
## load package
library("gplots")
```

Select subset of expression matrix corresponding to differentially expressed probes and plot heatmap. I was not able to reproduce the heatmap in the paper fully, and I think it has to do with how they cluster the probes.

```{r}
exprs.data <- exprs(macrophage.data)[rownames(exprs(macrophage.data)) %in% topProbes2, ]

## plot heatmap
a <- heatmap.2(exprs.data, key=TRUE, scale="row", density.info="none", trace="none", labRow="", Colv=FALSE, Rowv=FALSE, dendrogram="none", labCol=pData(macrophage.data)$ptid) 

## labeling for heatmap
mtext("Mock", side=3, line=-4, at=0.37); mtext("|", side=3, line=-4, at=0.55); mtext("Poly I:C", side=3, line=-4, at=0.73)
mtext("VL-", side=3, line=-5, at=0.28); mtext("|", side=3, line=-5, at=0.37); mtext("VL+", side=3, line=-5, at=0.46)
mtext("VL-", side=3, line=-5, at=0.63); mtext("|", side=3, line=-5, at=0.75); mtext("VL+", side=3, line=-5, at=0.82)
```

